Efficient forwards-time simulation with ancestral recombination graphs

### 2017-11-03 forwards_paper.tex
Added some text on mutations.

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10. In the
        left-most panel we show the tree sequence pictorially in two different ways:
        in the top are the tree topologies and the bottom shows the spatial extent of the
        edges that define these topologies. The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
        by \msprime.
        \jk{The figure needs to be rescaled a bit to give more space to the trees. We could also use a
        time scale next to the trees.}
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
A mutation occurs over a specific node at a given site, and results in a specific
derived state. Thus, all samples below the mutation node in the tree will inherit
this state (unless further mutations are encountered). Three mutations are shown
in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
mutation occurs at site zero (which is in the left-hand tree), which is a simple
mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
right hand tree) has two mutations: one occurring over node $3$ changing the state to
`C' and a back mutation over node $1$ changing the state to `G'.
\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
Perhaps if we added the state changes to the trees, and the states of the leaves
below the samples, we could get rid of all the example discussion here?}

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million tree
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB. Working with either the Newick or VCF encoding
of this dataset would be exceedingly cumbersome, likely requiring several
days of CPU time simply to read the information info memory. \jk{Could
add concrete timing estimates based on subsets here, but maybe overdoing it?}

The tree sequence encoding is very compact but it is also very efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}

@@

\begin{enumerate}
\item Explain how to add mutational information. We can also do this afterwards on the
output trees if they are nuetral.
\item Explain that we can input these trees into an mpsrime tree sequence and they will
work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
instead though.
\item However, there is huge amount of redundancy in these. The amount of information stored
grows linearly with time.

>>>>

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10. In the
        left-most panel we show the tree sequence pictorially in two different ways:
        in the top are the tree topologies and the bottom shows the spatial extent of the
        edges that define these topologies. The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
        by \msprime.
        \jk{Some minor alignment issues present here.}
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
A mutation occurs over a specific node at a given site, and results in a specific
derived state. Thus, all samples below the mutation node in the tree will inherit
this state (unless further mutations are encountered). Three mutations are shown
in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
mutation occurs at site zero (which is in the left-hand tree), which is a simple
mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
right hand tree) has two mutations: one occurring over node $3$ changing the state to
`C' and a back mutation over node $1$ changing the state to `G'.

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million tree
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB. Working with either the Newick or VCF encoding
of this dataset would be exceedingly cumbersome, likely requiring several
days of CPU time simply to read the information info memory.

The tree sequence encoding is compact but it is also efficient to process.
Many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies. For example, the
pairwise nucleotide diversity $\pi$, is defined as the average number of
differences between sequences in the sample. Computing $\pi$ directly
from observed sequence data requires $O(n^2 m)$ time, since we must
compare all pairs of samples at every site. However, given the topology of
the trees at every site and the locations of mutations much better algorithms
are possible. Using the fact that we can count the number of samples below
a given node efficiently using previously described tree sequence
algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
becomes roughly $O(n \log n + m)$.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, and calculating $\pi$ over all 500K samples
in the example above required about 1.2 seconds.
\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
it too deeply. I guess we should also compare this with something else in order to
show that this time is pretty good. We could get a subset of the sites and push
the data into pylibseq maybe, or scikit-allel??}

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that infinite sites mutations
occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a discinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed. In the case of infinite sites mutations
there is no difference between generating these mutations during the forwards
in time simulation and after it has completed. We are using precisely the same
information.

It is significantly more efficient to generate mutations as a separate process
after the topological simulation has completed because we will generate many
mutations that will be lost in the population.

\begin{enumerate}
\item Explain that we can input these trees into an mpsrime tree sequence and they will
work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
instead though.
\item However, there is huge amount of redundancy in these. The amount of information stored
grows linearly with time.
